NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-04-24 Manuscript/manuscriptPS.tex
Small updates

The molecular level interpretation of these observations is,
however, lacking, and  classical molecular dynamics (MD) simulations have been widely used in efforts to
to understand the PS headgroup structure, its influence on lipid bilayer properties, and its
interaction with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
Unfortunately, the results have depended on the force field used.
For example, recent simulations using the NBfix parameters for calcium \cite{kim16} in
CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}; in contrast,
the results from same lipid model without the NBfix ion parameters, combined with NMR chemical shifts and
rotational-echo double-resonance (REDOR) experiments, indicate a significant binding affinity also to the phosphate region \cite{hallock18}.
Meanwhile, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
combined with fluorescent and vibrational sum frequency spectroscopy, suggest substantial
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.

We have recently demonstrated that the lipid C--H bond order parameters, $S_\mathrm{CH}$~\cite{botan15,catte16}, can be utilized to resolve such controversies. The $S_\mathrm{CH}$ can be
measured from NMR experiments with high accuracy and directly compared to simulations
in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}. Using this approach,
it established that the structure of PC lipid headgroup and glycerol backbone are not well
captured by most MD force fields~\cite{botan15}, and that the cation binding to PC
lipid bilayers is overestimated \cite{catte16}.
%
%Based on these data, the cation binding affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
%\todo{Should we leave the mention of ECC out from the Introduction
%(i.e., mention it only in the Conclusions) as the ECC parameters are not used in the paper?}

@@

\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPSformatted.pdf}
  \caption{\label{NAdensPOPS}
    Counterion density profiles along the membrane normal and areas per lipid (A$_{\rm pl}$)
    of POPS lipid bilayer from simulations with different force fields.
    The experimental area per lipid is~62.7~$\pm$~1.3~\AA$^2$~\cite{pan14}.
  }
  \todo{Commented by M. Javanainen in blog:
    MacRog pure POPS is simulated with Verlet cutoff scheme, Piggot is rerunning with group cutoff scheme. Check if affects results \& update figures when ready} \\
  \todo{It has been suggested that we should format this with the same style as in NMRlipids II: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/44}
\end{figure}

@@

\subsection{Ca$^{2+}$ binding affinity to bilayers with negatively charged PS lipids}
\begin{figure}[tb]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixtureLOWconsformatted.pdf}
  \caption{\label{CAdensPCPSmixture}
    Number density profiles of Ca$^{2+}$ from POPC:POPS (5:1) mixtures simulated with different force fields.
    The smallest simulated CaCl$_2$ concentrations are shown.
    The density profiles for all the simulated concentrations are given in SI figure \ref{CAdensPCPSmixtureALL}.
  }
  \todo{Should we include also counterions into the plot?} \\
  \todo{It has been suggested that we should format this with the same style as in NMRlipids II: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/44}
\end{figure}

@@

Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
exhibit a strong dependence of CaCl$_2$ with small concentrations that saturates below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order paramter increases with added CaCl$_2$,
whereas the larger $\alpha$-carbon order parameter decreases; a slight increase is observed in
the smaller $\alpha$-carbon value. All these changes are significantly overestimated in the
tested simulation models, including CHARMM36 which underestimated the binding affinity.
Importantly, the different simulation models predict qualitatively different behaviour
for the two POPS $\alpha$-carbon order parameters with added calcium.
For example, both order parameters decrease in Berger, but increase
in MacRog, and in Lipid14/17 and CHARMM36 a more complicated behavior is seen.
This is in contrast to the PC headgroup, where
qualitatively correct reponse to bound ions is observed
in all simulation models, despite significant discrepancies in the headgroup
structures produced in salt-free simulations~\cite{catte16}.
The divergent response of the Berger model may arise from the ring like structures
observed in the headgroup region in this model (Fig. 6 in Ref. \citenum{mukhopadhyay04}).
%These ring like structures are a widespread feature of typical Berger based lipid force
%fields containing explicit hydrogen atoms in the head group
%[REFS https://pubs.acs.org/doi/abs/10.1021/jp900645z and https://www.sciencedirect.com/science/article/pii/S0300908408000692 and https://pubs.acs.org/doi/abs/10.1021/jp9110019]
Therefore, we conclude that
improvement of force fields is necessary to correctly capture the interactions between the
PS headgroup and calcium ions in MD simulations.
\section{Conclusions}

@@

First, we complemented the available experimental information \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters, and the proceeded to compare MD simulation results from several force fields to the experimental data
This revealed that none of the force fields
%tested using the NMRlipids open collaboration
reproduce the PS headgroup order parameters within the experimental accuracy. However,
the best models for the serine headgroup suggested a characteristic rigid conformation for its
carboxyl group. Comparison to the experimentally observed order parameters in POPC:POPS (5:1) bilayers  at varying ion
concentrations \cite{roux90} then showed that the tested MD force fields
overestimate the cation binding affinity to these bilayers with two exceptions: 1) the MacRog simulation with potassium counterions appears to produce the most realistic monovalent ion binding
affinity to PS-containing lipid bilayers, and 2) the CHARMM36 force field with the recently introduced
NBfix correction for calcium \cite{kim16} underestimated the calcium binding affinity.
The experimentally measured response of the PS headgroup order parameter to the bound calcium, and to the dilution of bilayer with zwitterionic PC lipids, were not
qualitatively reproduced in any of the tested force fields, which underlines the need for more accurate
the MD force fields to study the biological functioning of PS lipids.
This is in contrast to the previous results with PC lipids,
where the experimentally measured headgroup order parameter responses to the bound charge
were in qualitative agreement even though the headgroup structures themselves were
incorrect and the cation binding affinities overestimated \cite{catte16}.

>>>>

The molecular level interpretation of these observations is,
however, lacking, and  classical molecular dynamics (MD) simulations have been widely used in efforts to
to understand the PS headgroup structure, its influence on lipid bilayer properties, and its
interaction with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
Unfortunately, the results have depended on the force field used.
For example, recent simulations using the NBfix parameters for calcium \cite{kim16} in
CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}; in contrast,
the results from the same lipid model without the NBfix ion parameters, combined with NMR chemical shifts and
rotational-echo double-resonance (REDOR) experiments, indicate a significant binding affinity also to the phosphate region \cite{hallock18}.
Meanwhile, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
combined with fluorescent and vibrational sum frequency spectroscopy, suggest substantial
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.

We have recently demonstrated that the lipid C--H bond order parameters, $S_\mathrm{CH}$,
can be utilized to resolve such controversies~\cite{botan15,catte16}. The $S_\mathrm{CH}$ can be
measured from NMR experiments with high accuracy and directly compared to simulations
in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}. Using this approach,
it established that the structure of PC lipid headgroup and glycerol backbone are not well
captured by most MD force fields~\cite{botan15}, and that the cation binding to PC
lipid bilayers is overestimated \cite{catte16}.
%
%Based on these data, the cation binding affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
%\todo{Should we leave the mention of ECC out from the Introduction
%(i.e., mention it only in the Conclusions) as the ECC parameters are not used in the paper?}

@@

\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPSformatted.pdf}
  \caption{\label{NAdensPOPS}
    Counterion density profiles along the membrane normal and areas per lipid (A$_{\rm pl}$)
    of POPS lipid bilayer from simulations with different force fields.
    The experimental area per lipid is~62.7~$\pm$~1.3~\AA$^2$~\cite{pan14}.
  }
  \todo{Commented by M. Javanainen in blog:
    MacRog pure POPS is simulated with Verlet cutoff scheme, Piggot is rerunning with group cutoff scheme. Check if affects results \& update figures when ready} \\
\end{figure}

@@

\subsection{Ca$^{2+}$ binding affinity to bilayers with negatively charged PS lipids}
\begin{figure}[tb]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixtureLOWconsformatted.pdf}
  \caption{\label{CAdensPCPSmixture}
    Number density profiles of Ca$^{2+}$ from POPC:POPS (5:1) mixtures simulated with different force fields.
    The smallest simulated CaCl$_2$ concentrations are shown.
    The density profiles for all the simulated concentrations are given in SI figure \ref{CAdensPCPSmixtureALL}.
  }
  \todo{Should we include also counterions into the plot?} \\
\end{figure}

@@

Experimentally, the POPS headgroup order parameters in POPC:POPS (5:1) mixtures
exhibit a strong dependence of CaCl$_2$ with small concentrations that saturates below 100 mM (Fig. \ref{changesWITHCaClPS}).
The $\beta$-carbon order parameter increases with added CaCl$_2$,
whereas the larger $\alpha$-carbon order parameter decreases; a slight increase is observed in
the smaller $\alpha$-carbon value. All these changes are significantly overestimated in the
tested simulation models, including CHARMM36 which underestimated the binding affinity.
Importantly, the different simulation models predict qualitatively different behaviour
for the two POPS $\alpha$-carbon order parameters with added calcium.
For example, both order parameters decrease in Berger, but increase
in MacRog, and in Lipid14/17 and CHARMM36 a more complicated behavior is seen.
This is in contrast to the PC headgroup, where
qualitatively correct reponse to bound ions is observed
in all simulation models, despite significant discrepancies in the headgroup
structures produced in salt-free simulations~\cite{catte16}.
The divergent response of the Berger model may arise from the ring like structures
observed in the headgroup region in this model (Fig. 6 in Ref. \citenum{mukhopadhyay04}).
%These ring like structures are a widespread feature of typical Berger based lipid force
%fields containing explicit hydrogen atoms in the head group
%[REFS https://pubs.acs.org/doi/abs/10.1021/jp900645z and https://www.sciencedirect.com/science/article/pii/S0300908408000692 and https://pubs.acs.org/doi/abs/10.1021/jp9110019]
Therefore, we conclude that
improvement of force fields is necessary to correctly capture the interactions between the
PS headgroup and calcium ions in MD simulations.
\section{Conclusions}

@@

First, we complemented the available experimental information \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters,
and then proceeded to compare MD simulation results from several force fields to the experimental data
This revealed that none of the force fields
%tested using the NMRlipids open collaboration
reproduce the PS headgroup order parameters within the experimental accuracy. However,
the best models for the serine headgroup suggested a characteristic rigid conformation for its
carboxyl group. Comparison to the experimentally observed order parameters in POPC:POPS (5:1) bilayers  at varying ion
concentrations \cite{roux90} then showed that the tested MD force fields
overestimate the cation binding affinity to these bilayers with two exceptions: 1) the MacRog simulation with potassium counterions appears to produce the most realistic monovalent ion binding
affinity to PS-containing lipid bilayers, and 2) the CHARMM36 force field with the recently introduced
NBfix correction for calcium \cite{kim16} underestimated the calcium binding affinity.
The experimentally measured response of the PS headgroup order parameter to the bound calcium, and to the dilution of bilayer with zwitterionic PC lipids, were not
qualitatively reproduced in any of the tested force fields, which underlines the need for more accurate
the MD force fields to study the biological functioning of PS lipids.
This is in contrast to the previous results with PC lipids,
where the experimentally measured headgroup order parameter responses to the bound charge
were in qualitative agreement even though the headgroup structures themselves were
incorrect and the cation binding affinities overestimated \cite{catte16}.
